Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-25 paper/current/main.tex
data availability, yy comments
fixed references
proofreading, main text

% Refs
\usepackage[style=nature,
					backend=biber,
					sortcites=true,
					autocite=superscript
]{biblatex}
\addbibresource{mobility.bib}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity are of paramount importance.
Here, using three discrete human trajectory datasets,  we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncovers cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, the embedding vectors reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has been constraining human movements throughout history, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or functional distances~\autocite{boschma2005proximity, brown1970migration, kim2018functional} from real-world data or \textit{a priori} relationships between geographic entities.

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.  
Low dimensional distance alone cannot represent the multitude of inter-related factors that drive mobility. 
Networks offer a solution to representing many dimensions of mobility, yet edges only encode simple relationships between connected entities.
Thus, capturing the complexity of mobility requires moving beyond simple functional distances and networks, to learning high-dimensional landscapes of mobility that incorporate the many facets of mobility into a single fine-grained and continuous representation.

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense and continuous vector-space representation, from which we can not only derive a meaningful functional distance between places but also probe relationships based on culture, language, and even prestige along with geographic relationship. 
We embed trajectories from three massive datasets: U.S. passenger flight itinerary records, South Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records.

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States. 
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports. 
The Korean accommodation reservations consist of accommodation reservation history between 2018 and 2020 between 1,038 unique accommodation locations in Seoul, South Korea.
A trajectory is constructed for each customer's accommodation reservation history, constructing an ordered sequence of accommodations reserved over time.
Finally, we use scientific mobility data that captures the affiliation trajectories of nearly 3 million scientists across ten years.
We focus in more detail on scientific mobility due to its richness and importance. 
Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, morgan2018prestige}, is not only an important topic in the Science of Science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, deville2014career}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

@@

A vector-space embedding of locations (airports and organizations) is learned by using trajectories as input to the standard skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered} and networks~\autocite{grover2016node2vec}.
The model is also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Construction of affiliation trajectories from publication records}
        \textbf{a.}
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
        A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication.
        \textbf{b.}
        If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

To validate our approach, we evaluate the quality of vector representations with their performance in predicting real-world mobility flows using the gravity model framework~\autocite{zipf1946gravity}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity} that models the \textit{expected flux}, $\hat{T}_{ij}$ between locations (airports and organizations) based on their populations and distance:

@@

where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods). 
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers on each airport, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, ($\hat{T}_{ij }= \hat{T}_{ji}$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

%
% Subsection: Embedding performance
%
% TODO: remove the subsection label?
%
\subsection*{Embeddings provide functional distance between locations}

We show that the embedding distance better predicts actual mobility flows than the geographic distance across three disparate datasets.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).  Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of South Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d). 
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c and Fig.~\ref{fig:supp:predict_geo_bymodel}) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d and  Fig.~\ref{fig:supp:predict_emb_bymodel}).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores  calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
In sum, our results demonstrate that, consistently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts U.S flights, accommodation reservations, and global scientific mobility.}
		\textbf{a} Embedding distance (cosine distance between organization vectors, top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom). 
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{b} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom). 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility. 
		For South Korean accommodation reservations, embedding distance better explains the expected flux than does geographic distance (\textbf{c}), and produces better predictions (\textbf{d}).
		Similarly, in the case of global scientific mobility, embedding distance explains the expected flux between organizations (\textbf{e)} and allows for better predictions (\textbf{f}) than geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}

@@

Globally, the geographic constraints are conspicuous; organizations tend to form clusters based on their national affiliations and national clusters tend to be near their geographic neighbors.
At the same time, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions much more clearly than alternative network representations (Fig.~\ref{fig:supp:network_vis}) that have been dominating studies of scientific mobility~\autocite{chinchilla2018global, czaika2018globalisation}.

The embedding space also allows us to \textit{zoom in} on subsets and re-project them to reveal local relationships.
For example, re-projecting organizations located in Western, Southern, and Southeastern Asia with UMAP (Fig.~\ref{fig:projection}b) reveals a gradient of countries between Egypt and the Philippines that largely corresponds to geography, but with some exceptions along with the cultural and religious spectra;
Malaysia, with its official religion of Islam, is nearer to Middle Eastern countries in the embedding space than many geographically-closer South Asian countries.
We also validate this finding quantitatively with the cosine distance between nations (the centroids of organizations vectors belonging to that country). 
Malaysia is nearer to many Islamic countries such as Iraq ($d = 0.29$), Pakistan ($d = 0.31$), and Saudi Arabia ($d = 0.41$)  than neighboring but Buddhist Thailand ($d = 0.43$) and predominantly-secular Vietnam ($d = 0.48$).

Linguistic and historical ties also affect scientific mobility.
We observe that Spanish-speaking Latin American nations are positioned near Spain (Fig.~\ref{fig:projection}c), rather than Portuguese-speaking Brazil ($d = 0.34$ vs. $d = 0.52$ for Mexico and $d = 0.39$ vs. $d = 0.49$ for Chile) reflecting linguistic and cultural ties.
Similarly, North-African countries that were once under French rule such as Morocco are closer to France ($d =0.35$) than to similarly geographically-distant European countries such as Spain ($d = 0.40$), Portugal ($d = 0.55$), and Italy ($d = 0.53$).
Comparable structural patterns exist even within a single country.
For example, organizations within Quebec in Canada are located nearer the French cluster ($d = 0365$) than the United States cluster ($d = 0.49$).

@@

%
% BIG PROJECTION
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Projection/big_embedding_picture.pdf}
	\caption{
		\textbf{Projection of embedding space reveals complex multi-scale structure of organizations.}
		\textbf{a.}
		UMAP projection \autocite{mcinnes2018umap} of the embedding space reveals country-level clustering.
		Each point corresponds to an organization and its size indicates the average annual number of mobile and non-mobile authors affiliated with that organization from 2008 to 2019.
		Color indicates the region.
		The separation of organizations in Quebec and the rest of Canada is highlighted. 
		\textbf{b.} Zooming into (re-projecting) the area containing countries in the Middle East and South and Southeast Asia shows a geographic and cultural gradient of country clusters.
		\textbf{c.} Similarly, zooming into the area organizations in Spain, Portugal, South, and Central America shows clustering by most widely-spoken majority language group--Spanish and Portuguese.
		\textbf{d.} Doing the same for organizations in the United States reveals geographic clustering based on state, roughly grouped by Census Bureau-designated regions,
		\textbf{e.} Zooming in further on Massachusetts reveals clustering based on urban center (Boston, Worcester), organizational sector (hospitals vs. universities), and university systems and prestige (UMass system vs. Harvard, MIT, etc.).
	}
	\label{fig:projection}
\end{figure}

Just as the embedding space makes it possible to \textit{zoom in} on subsets of organizations, it is also possible to \textit{zoom out}  by aggregating organizational vectors.
We can examine the country-level structure that governs scientific mobility.
For the purpose, we define the representative vector of each country as the average of their organizational vector and perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space based on their cosine similarity, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a).
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
The cluster structure shows that not only geography but also linguistic ties between countries are related to scientific mobility.

We quantify the relative importance of geography (by region), and language (by most widely-spoken language of each country) using the element-centric clustering similarity~\autocite{gates2019element}, a method that can compare hierarchical clustering by explicitly adjusting the relative importance of different levels of the hierarchy with scaling parameter $r$, acting like a \textit{zooming lens}.
If $r$ is high, the similarity is based on the lower levels of the dendrogram, whereas when $r$ is low, the similarity is based on higher levels.
Fig. \ref{fig:clustering}b demonstrates that regional relationships play a major role at higher levels of the clustering process (low $r$), and language (family) explains the clustering more at the lower levels (high $r$).
This suggests that the embedding space captures the hierarchical structure of mobility.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility, where researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Could the embedding space, to which no explicit prestige information is given, encode a prestige hierarchy?
This question is tested by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
As an external proxy of prestige, we use the Times Ranking of World Universities (we also use research impact from the Leiden Ranking~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the average vector of the top five U.S. universities according to the rankings, whereas the low-rank pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on the geometrical spectrum from the high-ranked to low-ranked poles (see Data and Methods).

>>>>

% Refs
\usepackage[style=nature,
					backend=biber,
					sortcites=true,
					autocite=superscript
]{biblatex}

% Ignore irrelevant biblatex fields
\AtEveryBibitem{%
 \clearfield{url}%
   \clearfield{month}%
 \clearfield{issn}%
 \clearfield{doi}%%
}

\addbibresource{mobility.bib}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, sugimoto2017mostimpact, petersen2018multiscale}.
Historically, human mobility has been strongly constrained by physical separation---geographic distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018globalisation, buchan2009globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important.
As understanding mobility is becoming critical for contemporary society, finding frameworks that can capture this complexity are of paramount importance.
Here, using three datasets of discrete human trajectories, we demonstrate that a neural embedding model~\autocite{mikolov2013word2vec} can encode nuanced relationships between locations into a vector-space, providing an effective measure of distance that reflects the multi-faceted structure of human mobility. 
Focusing on the case of scientific mobility, we show that embeddings of scientific organizations uncovers cultural and linguistic relations, and even academic prestige, at multiple levels of granularity.
Furthermore, the embedding vectors reveal universal relationships between organizational characteristics, and their place in the global landscape of scientific mobility. 
The ability to learn scalable, dense and meaningful representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has been constraining human movements throughout history, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or functional distances~\autocite{boschma2005proximity, brown1970migration, kim2018functional}, from real-world data or \textit{a priori} relationships between geographic entities.

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.  
Low dimensional distance alone cannot represent the multitude of inter-related factors that drive mobility. 
Networks offer a solution to representing many dimensions of mobility, yet edges only encode simple relationships between connected entities.
Capturing the complexity of mobility requires moving beyond simple functional distances and networks, to learning high-dimensional landscapes of mobility that incorporate the many facets of mobility into a single fine-grained and continuous representation.

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense and continuous vector-space representation, from which we can not only derive a meaningful functional distance between locations but also probe relationships based on culture, language, and even prestige along with geographic relationship. 
We embed trajectories from three massive datasets: U.S. passenger flight itinerary records, South Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records.

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States. 
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports. 
The South Korean accommodation reservations consist of customer reservation histories across 2018 and 2020 for 1,038 unique accommodation locations in Seoul, South Korea.
A trajectory is constructed for each customer, containing the ordered sequences of accommodations they reserved over time.
Finally, we use scientific mobility data that captures the affiliation trajectories of nearly 3 million scientists across ten years.
We focus in more detail on scientific mobility due to its richness and importance. 
Scientific mobility---which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, morgan2018prestige}---is not only an important topic in the Science of Science but also ideal for our study thanks to its well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, deville2014career}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

@@

A vector-space embedding of locations (airports, accommodations, and organizations) is learned by using trajectories as input to the standard with skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered} and networks~\autocite{grover2016node2vec}.
The model is also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Construction of affiliation trajectories from publication records.}
        \textbf{a.}
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
        A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication.
        \textbf{b.}
        If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

To validate our approach, we evaluate the quality of vector representations with their performance in predicting real-world mobility flows using the gravity model framework~\autocite{zipf1946gravity}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity} that models the \textit{expected flux}, $\hat{T}_{ij}$, between locations based on their populations and distance:

@@

where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods). 
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers on each airport, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, ($\hat{T}_{ij }= \hat{T}_{ji}$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}).

%
% Subsection: Embedding performance
%
%
\subsection*{Embeddings provide functional distance between locations}

We show that the embedding distance better predicts actual mobility flows than the geographic distance across three disparate datasets.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).  Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of South Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d). 
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d and  Fig.~\ref{fig:supp:predict_emb_bymodel}).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores  calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
In sum, our results demonstrate that, consistently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility in flights, accommodation reservations, and global scientific mobility.}
		\textbf{a} Embedding distance (cosine distance between organization vectors, top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom). 
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{b} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom). 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility. 
		For South Korean accommodation reservations, embedding distance better explains the expected flux than does geographic distance (\textbf{c}), and produces better predictions (\textbf{d}).
		Similarly, in the case of global scientific mobility, embedding distance explains the expected flux between organizations (\textbf{e)} and allows for better predictions (\textbf{f}) than geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}

@@

Globally, the geographic constraints are conspicuous; organizations tend to form clusters based on their national affiliations and national clusters tend to be near their geographic neighbors.
At the same time, the embedding space also reflects a mix of geographic, historic, cultural, and linguistic relationships between regions much more clearly than alternative network representations (Fig.~\ref{fig:supp:network_vis}) that have been common in studies of scientific mobility~\autocite{chinchilla2018global, czaika2018globalisation}.

The embedding space also allows us to \textit{zoom in} on subsets and re-project them to reveal local relationships.
For example, re-projecting organizations located in Western, Southern, and Southeastern Asia with UMAP (Fig.~\ref{fig:projection}b) reveals a gradient of countries between Egypt and the Philippines that largely corresponds to geography, but with some exceptions seemingly stemming from cultural and religious similarity;
Malaysia, with its official religion of Islam, is nearer to Middle Eastern countries in the embedding space than many geographically-closer South Asian countries.
We validate this finding quantitatively with the cosine distance between nations (the centroids of organizations vectors belonging to that country). 
Malaysia is nearer to many Islamic countries such as Iraq ($d = 0.29$), Pakistan ($d = 0.31$), and Saudi Arabia ($d = 0.41$)  than neighboring but Buddhist Thailand ($d = 0.43$) and predominantly-secular Vietnam ($d = 0.48$).

Linguistic and historical ties also affect scientific mobility.
We observe that Spanish-speaking Latin American nations are positioned near Spain (Fig.~\ref{fig:projection}c), rather than Portuguese-speaking Brazil ($d = 0.34$ vs. $d = 0.52$ for Mexico and $d = 0.39$ vs. $d = 0.49$ for Chile) reflecting linguistic and cultural ties.
Similarly, North-African countries that were once under French rule such as Morocco are closer to France ($d =0.35$) than to similarly geographically-distant European countries such as Spain ($d = 0.40$), Portugal ($d = 0.55$), and Italy ($d = 0.53$).
Comparable patterns exist even within a single country.
For example, organizations within Quebec in Canada are located nearer France ($d = 0.37$) than the United States ($d = 0.49$).

@@

%
% BIG PROJECTION
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Projection/big_embedding_picture.pdf}
	\caption{
		\textbf{Projection of embedding space reveals complex multi-scale structure of organizations.}
		\textbf{a.}
		UMAP projection \autocite{mcinnes2018umap} of the embedding space reveals country-level clustering.
		Each point corresponds to an organization and its size indicates the average annual number of mobile and non-mobile authors affiliated with that organization from 2008 to 2019.
		Color indicates the region.
		The separation of organizations in Quebec and the rest of Canada is highlighted. 
		\textbf{b.} Zooming into (re-projecting) the area containing countries in Western, South, and Southeast Asia shows a geographic and cultural gradient of country clusters.
		\textbf{c.} Similarly, zooming into the area containing organizations in Spain, Portugal, South, and Central America shows clustering by most widely-spoken majority language group: Spanish and Portuguese.
		\textbf{d.} Doing the same for organizations in the United States reveals geographic clustering based on state, roughly grouped by Census Bureau-designated regions,
		\textbf{e.} Zooming in further on Massachusetts reveals clustering based on urban center (Boston, Worcester), organizational sector (hospitals vs. universities), and university systems and prestige (UMass system vs. Harvard, MIT, etc.).
	}
	\label{fig:projection}
\end{figure}

Just as the embedding space makes it possible to \textit{zoom in} on subsets of organizations, it is also possible to \textit{zoom out}  by aggregating organizational vectors.
We can examine the country-level structure that governs scientific mobility.
For the purpose, we define the representative vector of each country as the average of their organizational vectors and, using their cosine similarities, perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a).
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
The cluster structure shows that not only geography but also linguistic ties between countries are related to scientific mobility.

We quantify the relative importance of geography (by region), and language (by the most widely-spoken language of each country) using the element-centric clustering similarity~\autocite{gates2019element}, a method that can compare hierarchical clustering by explicitly adjusting the relative importance of different levels of the hierarchy with a scaling parameter $r$, acting like a \textit{zooming lens}.
If $r$ is high, the similarity is based on the lower levels of the dendrogram, whereas when $r$ is low, the similarity is based on higher levels.
Fig. \ref{fig:clustering}b demonstrates that regional relationships play a major role at higher levels of the clustering process (low $r$), and language (family) explains the clustering more at the lower levels (high $r$).
This suggests that the embedding space captures the hierarchical structure of mobility.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility, in which researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Could the embedding space, to which no explicit prestige information is given, encode a prestige hierarchy?
This question is tested by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
As an external proxy of prestige, we use the Times Ranking of World Universities (we also use research impact from the Leiden Ranking~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the average vector of the top five U.S. universities according to the rankings, whereas the low-rank pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on the geometrical spectrum from the high-ranked to low-ranked poles (see Data and Methods).
